%% 摘要

\begin{abstract}
    低空无人机在巡检、测绘、导航和搜索等任务中具有主动运动、多视角观测和连续时空变化等特点。与普通图像问答或视频理解任务不同，无人机智能体不仅需要理解外部地标、场景和目标可见性，还需要理解自身相对位置、视角变化和飞行行为。围绕这一问题，本文开展面向无人机双认知时空推理的多模态大模型能力评测与优化方法研究，将无人机具身推理能力划分为自我状态认知和环境情况认知两条主线，并在图像与视频两种媒介下构建六类评测任务。

    在能力测试方面，本文实现了从 AerialVLN 仿真场景扫描、语义点云融合、地标资产复核、行为驱动视频生成到证据感知图像任务生成的四阶段自动化工具链，形成包含 12 个测试场景、512 个有效地标、4096 个图像样本和 2048 个视频样本的 UAV-DualCog 测试集。测试基准站点、代码仓库和数据集已公开发布。实验部分对闭源模型、开源模型和空间推理微调模型进行了统一评测，并从语义正确性、空间定位、时间定位、跨媒介迁移和双认知平衡等角度进行分析。结果表明，当前多模态大模型具备一定无人机场景理解能力，但仍普遍存在语义与证据脱节、自我状态认知弱于环境情况认知、跨媒介稳定性不足等问题。

    在能力优化方面，在优化方面，我们构建了包含 11232 个样本的 UAV-DualCog-Train 训练集，并针对 Qwen3.5-4B 模型进行了优化。训练流程采用了多模态低秩适配（LoRA）监督微调和联合群组相对策略优化（GRPO），以引导模型输出“答案+边界框”的结构化结果，从而增强语义与证据的一致性。通过实验结果可知，优化模型在图像任务上获得了显著直接效益，图像全局准确率提升 20.78\%，空间证据指标同步改善，在视频任务上全局准确率提升 9.91\%，证明图像侧证据感知训练能够产生部分跨媒介迁移。与此同时，可靠无人机多模态系统还要继续提高视频时序整合能力，解决图像空间证据和视频时间证据之间的跨媒介协同问题。
    \keywordslist{无人机, 多模态大模型, 时空推理, 测试基准, 监督微调, 群组相对策略优化}
\end{abstract}

\begin{engabstract}
    Low-altitude unmanned aerial vehicles operate under active motion, multi-view observation, and continuous spatio-temporal changes. Unlike conventional image question answering or video understanding, a UAV agent must understand not only external landmarks, scenes, and target visibility, but also its own relative position, viewpoint transition, and flight behavior. This thesis studies capability evaluation and optimization of multimodal large language models (MLLMs) for UAV dual-cognition spatio-temporal reasoning. It formulates embodied UAV reasoning along two dimensions, self-state cognition and environment-state cognition, and instantiates them through six image-based and video-based tasks.

    For evaluation, this thesis implements a four-stage construction toolchain covering AerialVLN scene scanning, semantic point-cloud fusion, landmark asset verification, behavior-driven video generation, and evidence-aware image task generation. The UAV-DualCog test set contains 12 scenes, 512 valid landmarks, 4,096 image samples, and 2,048 video samples. The benchmark website, code repository, and dataset have been publicly released. 30 mainstream proprietary models, open-weight models, and spatially specialized fine-tuned models are elvalueated under a unified protocol. The results show that current MLLMs have obtained partial UAV scene understanding ability, but still suffer from a gap between answers and evidence, weaker self-aware cognition than environment-aware cognition, and insufficient cross-media stability.

    For optimization, this thesis further designs and constructs UAV-DualCog-Train including 11232 samples to fine-tune Qwen3.5-4B using multimodal LoRA-based SFT and joint GRPO reward optimization. This encourages structured outputs with bounding boxes to enhance semantic-evidence consistency. Experiments show that the optimized model achieves clear direct gains on image tasks, with image global accuracy improved by 20.78\% points and spatial evidence metrics improved simultaneously. On video tasks, the global accuracy improves by 9.91\% points, indicating partial cross-media transfer from evidence-aware image training. However, reliable UAV multimodal systems still require stronger temporal integration and better coordination between spatial evidence in images and temporal evidence in videos.
    \engkeywordslist{UAV, multimodal large model, spatio-temporal reasoning, benchmark, SFT, GRPO}
\end{engabstract}
